|

### 2023-08-28 chapter_4/src/prr/dcr_color_and_non-political_behaviors_prr.qmd
doing a little bit more thinking on this chapter

While political polarization certainly reflects a growing gap and consolidation of two camps of policy preferences and attitudes [@dellaposta_pluralistic_2020;@luders_attitude_2023], it manifests as something more consequential for our every-day experiences. Political polarization has come to reflect social groups that distinguish "us" versus "them" with increasingly fewer and fewer non-political social groups that are not wrapped up in our partisan group [@mason_uncivil_2018]. That is, we are not only holding fewer and fewer cross-cutting attitudes that moderate our political viewpoints [@luders_attitude_2023], but we also have fewer and fewer cross-cutting social groups that encourage interaction and experiences with perspectives of out-partisans [@mason_uncivil_2018]. The consequence of such phenomena is that we are becoming increasingly willing to harbor strong negative feelings towards those of the other political party and strong positive affectations to members of our political party [@iyengar_affect_2012;@iyengar_fear_2015].

The consequences of this affective polarization encourages discomfort being around those who identify with the other political party and finding comfort around those that share the same partisanship. In fact, some evidence suggests that it is even more sinister than comfort but rather predicts lower empathy for outpartisans [@allamong_screw_2021], as well as an increased willingness to describe outpartisans as subhuman [@martherus_party_2021]. Racists commonly use dehumanizing language to describe racial minorities [@utych_how_2018].

Evidence of this discomfort and its manifestations with out-partisans is piling up. For example, @carlson_what_2022 demonstrates that individuals perceive that they may incur social costs with friends and neighbors by engaging in a conversation with those they disagree with and expressing that disagreement, and avoid such situations as a result. Other work corroborate this finding by suggesting that individuals will be less likely to share the social media posts of political viewpoints they agree with if they perceive that they are in the political minority of those that would see that sharing of the post [@van_der_does_strategic_2022]. As a result of this tendency to speak out as a political minority in front of a group of people, individuals have fewer outpartisans in their group of folks they regularly discuss politics with [@butters_polarized_2022].

@@

The consequences of the fewer cross-cutting social groups and increasingly homogenous cultural preferences of Republicans and Democrats manifest as dramatic differences in preferences about the communities that we are part of. @gimpel_seeking_2015 demonstrates that describing a neighborhood with explicit and implicit information (e.g., racial composition) about the partisan composition influences whether one expresses a desire or not to move and live there. Further work corroborates this finding by providing evidence from a number of studies which demonstrate that not only do people associate particular characteristics of a neighborhood such as the prevalence of churches with Republicans and hybrid cars with Democrats, but also that these features attract copartisans and repulse outpartisans [@motyl_how_2020].

Some work suggests, however, that these expressed preferences to live near copartisans do not carryover to migration patterns [see @iyengar_affect_2012]. In fact, there is some evidence suggesting that controlling for everything else but current rates of political sorting, we would have less political homogeneity than we had in 2008 [@martin_does_2020]. Further, evidence by @martin_does_2020 suggests that people switch their party registration after moving as opposed to moving to be more aligned with their party registration.

@@

As discussed thus far in the chapter, non-political objects are often associated with particular partisan leanings. For example, the modal car in a neighborhood conveys the partisan composition of the whole neighborhood [@hetherington_prius_2018;@gebru_using_2017]. In the same spirit as the rest of this book, we do not need to necessarily distinguish between whether someone drives a hybrid sedan and a Ford pickup truck. Information giving insight into one's partisanship may even be as subtle as their color choice in car. For a reader who thinks "hypothesizing about car color choice and partisanship is a reach", I agree. As stated at the outset of the chapter, the goal of the chapter is to extend past applications of the snap-judgment model to explicitly partisan objects such as yard signs as we did in Chapter 2 and to really challenge the model by looking at non-political objects that many never associate with partisanship.

@@

Applying the snap-judgment model to an individual's choice to move to a neighborhood yields the following hypotheses: with all else equal, Republicans that see that the modal color of car is Red are more likely to report wanting to move to the neighborhood than if the modal car color were blue ($H_1$); with all else equal, Democrats that see the modal color of Blue are more likely to report wanting to move to the neighborhood than if the modal car color were red ($H_2$); once we include more information about a neighborhood's partisan composition, then the strength of their desire to live in a neighborhood will increase if congruent with their partisan affiliation ($H_3$) and will weaken if incongruent with their partisan affiliation ($H_4$).

To test these hypotheses, I conduct an experiment involving a fictional real estate posting. The treatment involves randomly assigning participants to view one of two versions of a fictional real estate posting. The characteristics of the neighborhood are exactly the same between the two conditions except for the "favorite car" in the neighborhood. In the first condition the "favorite car" is a red SUV. In the second condition, the "favorite car" is the same make and model of SUV as in the first condition, but blue. Participants are then given a scale from 0-10 to rate the neighborhood and are presented with a Likert scale to indicate, "How much would [they] like to live in this neighborhood?" Once the participants have responded to these questions, they are then given a "second page" describing the neighborhood. The information subjects sees depends on which of two conditions they are randomly assigned into. In one condition, the information indicates that the neighborhood is "40 minutes from the nearest metro area" and "60% voted for the Republican candidate in the latest congressional election." The second condition presents the same information but instead describes the neighborhood as "10 minutes from the nearest metro area" and "60% voted for the Democratic candidate in the latest congressional election." With this new information, I ask participants "With this new information, please update your rating of the neighborhood" on the 0-10 scale and are again presented with a Likert scale to indicate, "How much would [they] like to live in this neighborhood?"

>>>>

While political polarization certainly reflects a growing gap and consolidation of two camps of policy preferences and attitudes [@dellaposta_pluralistic_2020;@luders_attitude_2023], it manifests as something more consequential for our every-day experiences. Political polarization has come to reflect social groups that distinguish "us" versus "them" with increasingly fewer and fewer non-political social groups that are not wrapped up in our partisan group [@mason_uncivil_2018]. That is, we are not only holding fewer and fewer cross-cutting attitudes that moderate our political viewpoints [@luders_attitude_2023], but we also have fewer and fewer cross-cutting social groups that encourage interaction and experiences with perspectives of out-partisans [@mason_uncivil_2018]. The consequence of such phenomena is that we are becoming increasingly willing to harbor strong negative feelings towards those of the other political party and strong positive affectations for members of our political party [@iyengar_affect_2012;@iyengar_fear_2015].

The consequences of this affective polarization encourages discomfort being around those who identify with the other political party and finding comfort around those that share the same partisanship. In fact, some evidence suggests that it is even more sinister than comfort but rather predicts lower empathy for out-partisans [@allamong_screw_2021], as well as an increased willingness to describe out-partisans as subhuman [@martherus_party_2021].

Evidence of this discomfort and its manifestations with out-partisans is piling up. For example, @carlson_what_2022 demonstrates that individuals perceive that they may incur social costs with friends and neighbors by engaging in a conversation with those they disagree with and expressing that disagreement, and avoid such situations as a result. Other work corroborates this finding by suggesting that individuals will be less likely to share the social media posts of political viewpoints they agree with if they perceive that they are in the political minority of those that would see that sharing of the post [@van_der_does_strategic_2022]. As a result of this tendency to speak out as a political minority in front of a group of people, individuals have fewer out-partisans in their group of folks they regularly discuss politics with [@butters_polarized_2022].

@@

The consequences of the fewer cross-cutting social groups and increasingly homogenous cultural preferences of Republicans and Democrats manifest as dramatic differences in preferences about the communities that we are part of. @gimpel_seeking_2015 demonstrates that describing a neighborhood with explicit and implicit information (e.g., racial composition) about the partisan composition influences whether one expresses a desire or not to move and live there. Further work corroborates this finding by providing evidence from a number of studies which demonstrate that not only do people associate particular characteristics of a neighborhood such as the prevalence of churches with Republicans and hybrid cars with Democrats, but also that these features attract co-partisans and repulse out-partisans [@motyl_how_2020].

Some work suggests, however, that these expressed preferences to live near co-partisans do not carryover to migration patterns [see @iyengar_affect_2012]. In fact, there is some evidence suggesting that controlling for everything else but current rates of political sorting, we would have less political homogeneity than we had in 2008 [@martin_does_2020]. Further, evidence by @martin_does_2020 suggests that people switch their party registration after moving as opposed to moving to be more aligned with their party registration.

@@

As discussed thus far in the chapter, non-political objects are often associated with particular partisan leanings. For example, the modal car in a neighborhood conveys the partisan composition of the whole neighborhood [@hetherington_prius_2018;@gebru_using_2017]. In the same spirit as the rest of this book, we do not need to necessarily distinguish between whether someone drives a hybrid sedan and a Ford pickup truck. Information giving insight into one's partisanship may even be as subtle as their color choice in car. For a reader who thinks "hypothesizing about car color choice and partisanship is a reach," I agree. As stated at the outset of the chapter, the goal of the chapter is to extend past applications of the snap-judgment model to explicitly partisan objects such as yard signs as we did in Chapter 2 and to really challenge the model by looking at non-political objects that many never associate with partisanship.

@@

Applying the snap-judgment model to an individual's choice to move to a neighborhood yields the following hypotheses: with all else equal, Republicans that see that the modal color of car is red are more likely to report wanting to move to the neighborhood than if the modal car color were blue ($H_1$); with all else equal, Democrats that see the modal color of blue are more likely to report wanting to move to the neighborhood than if the modal car color were red ($H_2$); once we include more information about a neighborhood's partisan composition, then the strength of their desire to live in a neighborhood will increase if congruent with their partisan affiliation ($H_3$) and will weaken if incongruent with their partisan affiliation ($H_4$).

To test these hypotheses, I conduct an experiment involving a fictional real estate posting. The treatment involves randomly assigning participants to view one of two versions of a fictional real estate posting. The characteristics of the neighborhood are exactly the same between the two conditions except for the "favorite car" in the neighborhood. In the first condition the "favorite car" is a red SUV. In the second condition, the "favorite car" is the same make and model of SUV as in the first condition, but blue. Participants are then given a scale from 0-10 to rate the neighborhood and are presented with a Likert scale to indicate, "How much would [they] like to live in this neighborhood?"

Once the participants have responded to these questions, they are then given a "second page" describing the neighborhood. The information subjects sees depends on which of two conditions they are randomly assigned into. In one condition, the information indicates that the neighborhood is "40 minutes from the nearest metro area" and "60% voted for the Republican candidate in the latest congressional election." The second condition presents the same information but instead describes the neighborhood as "10 minutes from the nearest metro area" and "60% voted for the Democratic candidate in the latest congressional election." With this new information, I ask participants "With this new information, please update your rating of the neighborhood" on the 0-10 scale and are again presented with a Likert scale to indicate, "How much would [they] like to live in this neighborhood?"
